Joint maximum-likelihood of phylogeny and ancestral states is not consistent

### 2018-10-01 appendix.tex
oops hat problem
code documentation and pub. quality figs
make plot the same size
rename xi subscripts

given $(\tau, t)$, there exist sets $\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t)$ such that $\xi_j\in\ancestralSplitPartition_j(\tau, t)$ satisfies

@@

\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t) &= \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV = \xi_j \mid \tau, t).

@@

Here, $\xi_j$ is one of possibly many equiprobable ancestral state splits in $\ancestralSplitPartition_j(\tau, t)$.

@@

    &= \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV = \xi_j \mid \tau, t)

@@

For each site split $\siteSplit\in\mathcal{P}(\siteSplitSet)$, we maximize the likelihood over all $\ancestralSplit\in\mathcal{P}(\ancestralSplitSet)$.
A maximum occurs at one of possibly several ancestral state splits in $\mathcal{P}(\ancestralSplitSet)$, defined via $\ancestralSplitPartition_j(\tau, t)$ for the $j$th site split.
As a simple example, say all branch lengths correspond to a probability $p$ ($< 1/2$) of changing character along that branch, with $t=\{p,p,p,p,p\}$.
The probabilities of observing ancestral state splits for $\siteSplit_1=\emptyset$ are
\[
\Pr(\siteSplitRV=\emptyset, \ancestralSplitRV=\emptyset \mid \tau, t) =
(1-p)^5,
\]
\[
\Pr(\siteSplitRV=\emptyset, \ancestralSplitRV=\{1\} \mid \tau, t) =
\Pr(\siteSplitRV=\emptyset, \ancestralSplitRV=\{2\} \mid \tau, t) =
p^3(1-p)^2,
\]
\[
\Pr(\siteSplitRV=\emptyset, \ancestralSplitRV=\{1,2\} \mid \tau, t) =
p^4(1-p).
\]
The set of most likely ancestral states contains a single element, here $\ancestralSplitPartition_1(\tau, t)=\{\emptyset\}$.
Then, taking $\xi_1\in\ancestralSplitPartition_1(\tau, t)$ we have
\[
\Pr(\siteSplitRV=\emptyset, \ancestralSplitRV=\xi_1 \mid \tau, t) =
\Pr(\siteSplitRV=\emptyset, \ancestralSplitRV=\emptyset \mid \tau, t) =
(1-p)^5.
\]
For $\siteSplit_5=\{1,2\}$ we have
\[
\Pr(\siteSplitRV=\{1,2\}, \ancestralSplitRV=\emptyset \mid \tau, t) =
\Pr(\siteSplitRV=\{1,2\}, \ancestralSplitRV=\{1,2\} \mid \tau, t) =
p^2(1-p)^3,
\]
\[
\Pr(\siteSplitRV=\{1,2\}, \ancestralSplitRV=\{1\} \mid \tau, t) =
\Pr(\siteSplitRV=\{1,2\}, \ancestralSplitRV=\{2\} \mid \tau, t) =
p^3(1-p)^2.
\]
Here, the set of most likely ancestral states is $\ancestralSplitPartition_5(\tau, t)=\{\emptyset,\{1,2\}\}$, and, for $\xi_5\in\ancestralSplitPartition_5(\tau, t)$,
\[
\Pr(\siteSplitRV=\{1,2\}, \ancestralSplitRV=\xi_5 \mid \tau, t) =
p^2(1-p)^3.
\]

@@

The likelihood in \eqref{eq:profile_likelihood} can be written as
\begin{align}
L_\nCols'(\tau, t; \fullAlignment) &= \max_{\fullAncestralStates} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates) \nonumber \\
                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ \Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) \nonumber \\
                             &\propto \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i), \ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \tau, t) \nonumber \\
                             &= \prod_{i=1}^{\nCols} \ \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t) \nonumber \\
                             &= \prod_{j=1}^{\nSiteSplits} \ \left[\Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
\end{align}
for $\siteSplit_j\in\mathcal{P}(\siteSplitSet)$ and some $\xi_j\in\ancestralSplitPartition_j(\tau, t)$ with $1 \le j \le q$ where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ that project to site split $\siteSplit_j$.

Let
\[
L_\nCols''(\tau, t; \fullAlignment) = \prod_{j=1}^{\nSiteSplits} \ \left[\Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t)\right] ^{\nCols_j(\fullAlignment)}
\]
be the final product in \eqref{eq:site_pattern_likelihood}.
Assume $\nCols$ observations are generated from a model with parameters $(\tau^*, t^*)$.
We have
\begin{equation*}
\frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t)
\end{equation*}
so that, in the $\nCols\rightarrow\infty$ limit,
\begin{equation}
\begin{split}
&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) \\
&\qquad \rightarrow \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t). \label{eq:site_pattern_profile_likelihood_mean}
\end{split}
\end{equation}

@@

%das: @EM I think this is actually the same Wald reference that Felsenstein tried to use.
%It works here for a fixed topology, and didn't work for him or your reference because I think the upper-semicontinuous condition doesn't hold for trees in general.
%Also I don't think the compactness condition holds---tree space isn't compact is it?
%However we now have a potential snag in that we are doing what Yang (1994) in his intro says is bad, i.e., doing ML for fixed, separate topologies and comparing them.
%We still have the behavior in the plots and the proofs, since they all assumed a fixed topology, and I think everything's fine with the arguments about Fels vs InvFels, but maybe you can let me know what you think?
For a fixed $\tau$, we show that $\hat{t}_\nCols \rightarrow \hat{t}$ for
\[
\hat{t}_\nCols = \arg\max_{t\in\mathcal{T}} \ \frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment)
\]
and
\[
\hat{t} = \arg\max_{t\in\mathcal{T}} \ \ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi).
\]
Using the notation in Section 5.2.1 in \citet{van1998asymptotic}, we let
\[
m_t(\alignmentColumn) = \sum_{j=1}^\nSiteSplits 1\{\patternToSplit(\alignmentColumn)=\siteSplit_j\} \cdot \log \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t)
\]
so that
\[
\frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment) = \frac{1}{n} \sum_{i=1}^\nCols m_t(\alignmentColumn_i)
\]
and
\[
\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi) = E[m_t].
\]
To show $\hat{t}_\nCols \rightarrow \hat{t}$, we use Wald's consistency proof \citep[p. 48, Theorem 5.14 of ][]{van1998asymptotic}, which requires four conditions.
The first is that $\mathcal{T}$ is compact, which is obviously true.
The second is that
\[
E\left[\sup_{t\in\mathcal{T}} m_t\right] < \infty,
\]
and, since $m_t(\alignmentColumn)$ is nonpositive for all $t$ and $\alignmentColumn$, this property holds.
The remaining conditions are on the maps
\[
\alignmentColumn \mapsto \sup_t m_t(\alignmentColumn)
\]
and
\[
t \mapsto m_t(\alignmentColumn).
\]
We need the first map to be measurable, which is evident since the domain $\alphabet^\nSiteRows$ of the mapping is a finite set, and so all subsets of the domain are also finite and thus measurable.
Finally, we must have the the second mapping be upper-semicontinuous for almost all $\alignmentColumn$.
For a fixed ancestral state split $t \mapsto m_t(\alignmentColumn)$ is continuous for all $\alignmentColumn$.
If we move about in $\mathcal{T}$, a different ancestral state split becomes more likely, though when we maximize over ancestral state splits we obtain a continuous function since the maximum over continuous functions is also continuous.
This ensures the upper-semicontinuous property of this mapping, and shows $\hat{t}_\nCols \rightarrow \hat{t}$, allowing our consistency results to be proved using $\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi)$.

@@

\begin{table}
\centering
\begin{tabular}{|lll|l|}
\hline
$\siteSplit_j$ & $\ancestralSplitPartition_j(\tau_1, t)$ & $\xi_j$ & $32\cdot\Pr(\siteSplitRV=\siteSplit_j,\ancestralSplitRV=\xi_j \mid \tau_1,t)$\\
\hline
$\emptyset$&$\{\emptyset\}$&$\emptyset$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1+w)$\\

@@

$\{1,3\}$  &$\{\emptyset,\{1\},\{1,2\}\}$&$\emptyset$&$(1-x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
&&$\{1\}$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1-w)$\\
&&$\{1,2\}$&$(1+x_1)(1-y_1)(1+x_2)(1-y_2)(1+w)$\\
\hline
\end{tabular}
\caption{
32 times likelihood values on the InvFels tree $\tau_1$.
Due to the symmetry of the likelihood, WLOG we let $x_2 \ge x_1$ and $y_2 \ge y_1$ and maximize over ancestral state splits to reduce the number of possible functional forms to consider.
Likelihoods with multiple entries have maxima determined by unknown branch length parameters.
Because in 4 cases there are 3 possibilities for $\xi_j$, there are $3^4=81$ possible forms for the likelihood.
}
\label{tab:likelihoods}
\end{table}

@@

\begin{table}
\centering
\begin{tabular}{|lll|l|}
\hline
$\siteSplit_j$ & $\ancestralSplitPartition_j(\tau_1, t)$ & $\xi_j$ & $32\cdot\Pr(\siteSplitRV=\siteSplit_j,\ancestralSplitRV=\xi_j \mid \tau_1,t)$\\
\hline
$\emptyset$&$\{\emptyset\}$&$\emptyset$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1+w)$\\
$\{1\}$    &$\{\emptyset\}$&$\emptyset$&$(1-x_1)(1+y_1)(1+x_2)(1+y_2)(1+w)$\\
$\{2\}$    &$\{\emptyset\}$&$\emptyset$&$(1+x_1)(1-y_1)(1+x_2)(1+y_2)(1+w)$\\
$\{3\}$    &$\{\emptyset,\{1\},\{1,2\}\}$&$\emptyset$&$(1+x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
$\{1,2,3\}$&$\{\emptyset,\{1\},\{1,2\}\}$&$\emptyset$&$(1-x_1)(1-y_1)(1-x_2)(1+y_2)(1+w)$\\
$\{1,2\}$  &$\{\emptyset\}$&$\emptyset$&$(1-x_1)(1-y_1)(1+x_2)(1+y_2)(1+w)$\\
$\{2,3\}$  &$\{\emptyset,\{1\},\{1,2\}\}$&$\emptyset$&$(1+x_1)(1-y_1)(1-x_2)(1+y_2)(1+w)$\\
$\{1,3\}$  &$\{\emptyset,\{1\},\{1,2\}\}$&$\emptyset$&$(1-x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
\hline
\end{tabular}
\caption{
32 times the maximal likelihood values on the InvFels tree $\tau_1$ where $\emptyset$ is the most likely ancestral state split for each site split.
}
\label{tab:likelihoods-restricted}
\end{table}

@@

\begin{figure}
    \begin{minipage}{0.95\linewidth}
        \centering
        \includegraphics[width=.95\textwidth]{analytic-anc-state}
        \label{ }
    \end{minipage}
    \begin{minipage}{0.95\linewidth}
        \centering
        \begin{tabular}{ll}
        \hline
        $\hat{\boldsymbol\xi_1}$ & $\{\emptyset, \emptyset, \emptyset, \emptyset, \emptyset, \emptyset, \emptyset, \emptyset\}$\\
        $\hat{\boldsymbol\xi_2}$ & $\{\emptyset, \emptyset, \emptyset, \emptyset, \{1,2\}, \emptyset, \emptyset, \emptyset\}$\\
        $\hat{\boldsymbol\xi_3}$ & $\{\emptyset, \emptyset, \emptyset, \emptyset, \{1,2\}, \emptyset, \emptyset, \{1\}\}$\\
        $\hat{\boldsymbol\xi_4}$ & $\{\emptyset, \emptyset, \emptyset, \emptyset, \{1,2\}, \emptyset, \{1,2\}, \{1\}\}$\\
        $\hat{\boldsymbol\xi_5}$ & $\{\emptyset, \emptyset, \emptyset, \{1\}, \{1\}, \emptyset, \{1\}, \{1\}\}$
        \end{tabular}
    \end{minipage}
\caption{
Regions of maximal ancestral state splits on the InvFels tree $\tau_1$.
}
\label{fig:max-anc-state}
\end{figure}

@@

The regions in Fig.~\ref{fig:max-anc-state} are analytically-derived regions of inconsistency in terms of probabilities of a character change along a branch for ``perfect'' data generated on the InvFels topology (Fig.~\ref{fig:farris-fels-top}) with $p_{w^*} = p_{y^*}$ (in terms of fidelities, $w^*=y^*$).
As the region of degeneracy in Fig.~\ref{fig:max-anc-state} gives the values of $x^*$ and $y^*$ where $\hat{w}$ is guaranteed to be one, we converge on a multifurcating topology in these cases.
%EM "letting $\emptyset$ be the maximal ancestral state split" sounds a little funny-- either it is or it isn't maximal.
It is easy to see that when $\emptyset$ is the maximal ancestral state split, we have the same log-likelihood for $\tau_1$ and $\tau_2$.
Moreover, if $w=1$, the internal branch becomes zero-length and the two topologies are indistinguishable.
Let $\mathcal{T}_0$ be such that, for $t^*=\{x^*,y^*,x^*,y^*,y^*\}$, $t^*\in\mathcal{T}_0$ corresponds to $x^*$ and $y^*$ falling in the region in Fig.~\ref{fig:max-anc-state} where $\hat{\boldsymbol\xi}=\hat{\boldsymbol\xi}_1$.
We can see this results in the likelihood of both topologies being equal, i.e.,
\begin{align*}
\max_{t : t^*\in\mathcal{T}_0} & \ \ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi)\\
&= \max_{t: \boldsymbol\xi=\hat{\boldsymbol\xi}_1, w=1, \tau=\tau_1} \ \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, \{x_1,y_1,x_2,y_2,w\})\\
&= \max_{t: \boldsymbol\xi=\hat{\boldsymbol\xi}_1, w=1, \tau=\tau_2} \ \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, \{x_1,y_1,x_2,y_2,w\}).
\end{align*}

>>>>

given $(\tau, t)$, there exist sets $\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t)$ such that $\xi_{\siteSplit_j}\in\ancestralSplitPartition_j(\tau, t)$ satisfies

@@

\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t) &= \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV = \xi_{\siteSplit_j} \mid \tau, t).

@@

Here, $\xi_{\siteSplit_j}$ is one of possibly many equiprobable ancestral state splits in $\ancestralSplitPartition_j(\tau, t)$.

@@

    &= \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV = \xi_{\siteSplit_j} \mid \tau, t)

@@

For each site split $\siteSplit\in\mathcal{P}(\siteSplitSet)$, we maximize the likelihood over all $\ancestralSplit\in\mathcal{P}(\ancestralSplitSet)$.
A maximum occurs at one of possibly several ancestral state splits in $\mathcal{P}(\ancestralSplitSet)$, defined via $\ancestralSplitPartition_j(\tau, t)$ for the $j$th site split.
As a simple example, say all branch lengths correspond to a probability $p$ ($< 1/2$) of changing character along that branch, with $t=\{p,p,p,p,p\}$.
The probabilities of observing ancestral state splits for $\siteSplit_1=\emptyset$ are
\[
\Pr(\siteSplitRV=\emptyset, \ancestralSplitRV=\emptyset \mid \tau, t) =
(1-p)^5,
\]
\[
\Pr(\siteSplitRV=\emptyset, \ancestralSplitRV=\{1\} \mid \tau, t) =
\Pr(\siteSplitRV=\emptyset, \ancestralSplitRV=\{2\} \mid \tau, t) =
p^3(1-p)^2,
\]
\[
\Pr(\siteSplitRV=\emptyset, \ancestralSplitRV=\{1,2\} \mid \tau, t) =
p^4(1-p).
\]
The set of most likely ancestral states contains a single element, here $\ancestralSplitPartition_1(\tau, t)=\{\emptyset\}$.
Then, taking $\xi_{\emptyset}\in\ancestralSplitPartition_1(\tau, t)$ we have
\[
\Pr(\siteSplitRV=\emptyset, \ancestralSplitRV=\xi_{\emptyset} \mid \tau, t) =
\Pr(\siteSplitRV=\emptyset, \ancestralSplitRV=\emptyset \mid \tau, t) =
(1-p)^5.
\]
For $\siteSplit_5=\{1,2\}$ we have
\[
\Pr(\siteSplitRV=\{1,2\}, \ancestralSplitRV=\emptyset \mid \tau, t) =
\Pr(\siteSplitRV=\{1,2\}, \ancestralSplitRV=\{1,2\} \mid \tau, t) =
p^2(1-p)^3,
\]
\[
\Pr(\siteSplitRV=\{1,2\}, \ancestralSplitRV=\{1\} \mid \tau, t) =
\Pr(\siteSplitRV=\{1,2\}, \ancestralSplitRV=\{2\} \mid \tau, t) =
p^3(1-p)^2.
\]
Here, the set of most likely ancestral states is $\ancestralSplitPartition_5(\tau, t)=\{\emptyset,\{1,2\}\}$, and, for $\xi_{12}\in\ancestralSplitPartition_5(\tau, t)$,
\[
\Pr(\siteSplitRV=\{1,2\}, \ancestralSplitRV=\xi_{12} \mid \tau, t) =
p^2(1-p)^3.
\]

@@

The likelihood in \eqref{eq:profile_likelihood} can be written as
\begin{align}
L_\nCols'(\tau, t; \fullAlignment) &= \max_{\fullAncestralStates} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates) \nonumber \\
                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ \Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) \nonumber \\
                             &\propto \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i), \ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \tau, t) \nonumber \\
                             &= \prod_{i=1}^{\nCols} \ \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, t) \nonumber \\
                             &= \prod_{j=1}^{\nSiteSplits} \ \left[\Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
\end{align}
for $\siteSplit_j\in\mathcal{P}(\siteSplitSet)$ and some $\xi_{\siteSplit_j}\in\ancestralSplitPartition_j(\tau, t)$ with $1 \le j \le q$ where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ that project to site split $\siteSplit_j$.

Let
\[
L_\nCols''(\tau, t; \fullAlignment) = \prod_{j=1}^{\nSiteSplits} \ \left[\Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, t)\right] ^{\nCols_j(\fullAlignment)}
\]
be the final product in \eqref{eq:site_pattern_likelihood}.
Assume $\nCols$ observations are generated from a model with parameters $(\tau^*, t^*)$.
We have
\begin{equation*}
\frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, t)
\end{equation*}
so that, in the $\nCols\rightarrow\infty$ limit,
\begin{equation}
\begin{split}
&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) \\
&\qquad \rightarrow \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, t). \label{eq:site_pattern_profile_likelihood_mean}
\end{split}
\end{equation}

@@

%das: @EM I think this is actually the same Wald reference that Felsenstein tried to use.
%It works here for a fixed topology, and didn't work for him or your reference because I think the upper-semicontinuous condition doesn't hold for trees in general.
%Also I don't think the compactness condition holds---tree space isn't compact is it?
%However we now have a potential snag in that we are doing what Yang (1994) in his intro says is bad, i.e., doing ML for fixed, separate topologies and comparing them.
%We still have the behavior in the plots and the proofs, since they all assumed a fixed topology, and I think everything's fine with the arguments about Fels vs InvFels, but maybe you can let me know what you think?
For a fixed $\tau$, we show that $\hat{t}_\nCols \rightarrow \hat{t}$ for
\[
\hat{t}_\nCols = \arg\max_{t\in\mathcal{T}} \ \frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment)
\]
and
\[
\hat{t} = \arg\max_{t\in\mathcal{T}} \ \ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi).
\]
Using the notation in Section 5.2.1 in \citet{van1998asymptotic}, we let
\[
m_t(\alignmentColumn) = \sum_{j=1}^\nSiteSplits 1\{\patternToSplit(\alignmentColumn)=\siteSplit_j\} \cdot \log \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, t)
\]
so that
\[
\frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment) = \frac{1}{n} \sum_{i=1}^\nCols m_t(\alignmentColumn_i)
\]
and
\[
\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi) = E[m_t].
\]
To show $\hat{t}_\nCols \rightarrow \hat{t}$, we use Wald's consistency proof \citep[p. 48, Theorem 5.14 of ][]{van1998asymptotic}, which requires four conditions.
The first is that $\mathcal{T}$ is compact, which is obviously true.
The second is that
\[
E\left[\sup_{t\in\mathcal{T}} m_t\right] < \infty,
\]
and, since $m_t(\alignmentColumn)$ is nonpositive for all $t$ and $\alignmentColumn$, this property holds.
The remaining conditions are on the maps
\[
\alignmentColumn \mapsto \sup_t m_t(\alignmentColumn)
\]
and
\[
t \mapsto m_t(\alignmentColumn).
\]
We need the first map to be measurable, which is evident since the domain $\alphabet^\nSiteRows$ of the mapping is a finite set, and so all subsets of the domain are also finite and thus measurable.
Finally, we must have the the second mapping be upper-semicontinuous for almost all $\alignmentColumn$.
For a fixed ancestral state split $t \mapsto m_t(\alignmentColumn)$ is continuous for all $\alignmentColumn$.
If we move about in $\mathcal{T}$, a different ancestral state split becomes more likely, though when we maximize over ancestral state splits we obtain a continuous function since the maximum over continuous functions is also continuous.
This ensures the upper-semicontinuous property of this mapping, and shows $\hat{t}_\nCols \rightarrow \hat{t}$, allowing our consistency results to be proved using $\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi)$.

@@

\begin{table}
\centering
\begin{tabular}{|lll|l|}
\hline
$\siteSplit_j$ & $\ancestralSplitPartition_j(\tau_1, t)$ & $\xi_{\siteSplit_j}$ & $32\cdot\Pr(\siteSplitRV=\siteSplit_j,\ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau_1,t)$\\
\hline
$\emptyset$&$\{\emptyset\}$&$\emptyset$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1+w)$\\

@@

$\{1,3\}$  &$\{\emptyset,\{1\},\{1,2\}\}$&$\emptyset$&$(1-x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
&&$\{1\}$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1-w)$\\
&&$\{1,2\}$&$(1+x_1)(1-y_1)(1+x_2)(1-y_2)(1+w)$\\
\hline
\end{tabular}
\caption{
32 times likelihood values on the InvFels tree $\tau_1$.
Due to the symmetry of the likelihood, WLOG we let $x_2 \ge x_1$ and $y_2 \ge y_1$ and maximize over ancestral state splits to reduce the number of possible functional forms to consider.
Likelihoods with multiple entries have maxima determined by unknown branch length parameters.
Because in 4 cases there are 3 possibilities for $\xi_{\siteSplit_j}$, there are $3^4=81$ possible forms for the likelihood.
}
\label{tab:likelihoods}
\end{table}

@@

\begin{table}
\centering
\begin{tabular}{|lll|l|}
\hline
$\siteSplit_j$ & $\ancestralSplitPartition_j(\tau_1, t)$ & $\xi_{\siteSplit_j}$ & $32\cdot\Pr(\siteSplitRV=\siteSplit_j,\ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau_1,t)$\\
\hline
$\emptyset$&$\{\emptyset\}$&$\emptyset$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1+w)$\\
$\{1\}$    &$\{\emptyset\}$&$\emptyset$&$(1-x_1)(1+y_1)(1+x_2)(1+y_2)(1+w)$\\
$\{2\}$    &$\{\emptyset\}$&$\emptyset$&$(1+x_1)(1-y_1)(1+x_2)(1+y_2)(1+w)$\\
$\{3\}$    &$\{\emptyset,\{1\},\{1,2\}\}$&$\emptyset$&$(1+x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
$\{1,2,3\}$&$\{\emptyset,\{1\},\{1,2\}\}$&$\emptyset$&$(1-x_1)(1-y_1)(1-x_2)(1+y_2)(1+w)$\\
$\{1,2\}$  &$\{\emptyset\}$&$\emptyset$&$(1-x_1)(1-y_1)(1+x_2)(1+y_2)(1+w)$\\
$\{2,3\}$  &$\{\emptyset,\{1\},\{1,2\}\}$&$\emptyset$&$(1+x_1)(1-y_1)(1-x_2)(1+y_2)(1+w)$\\
$\{1,3\}$  &$\{\emptyset,\{1\},\{1,2\}\}$&$\emptyset$&$(1-x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
\hline
\end{tabular}
\caption{
32 times the maximal likelihood values on the InvFels tree $\tau_1$ where $\emptyset$ is the most likely ancestral state split for each site split.
}
\label{tab:likelihoods-restricted}
\end{table}

@@

\begin{figure}
    \begin{minipage}{\linewidth}
        \centering
        \includegraphics[width=.95\textwidth]{analytic-anc-state}
        \label{ }
    \end{minipage}
    \begin{minipage}{\linewidth}
        \centering
        Maximal ancestral state split definitions
        \begin{tabular}{l}
        \hline
        $\hat{\boldsymbol\xi}=\{\xi_{\emptyset}, \xi_{1}, \xi_{2}, \xi_{3}, \xi_{123}, \xi_{12}, \xi_{23}, \xi_{13}\}$\\
        \hline
        $\hat{\boldsymbol\xi}_1=\{\emptyset, \emptyset, \emptyset, \emptyset, \emptyset, \emptyset, \emptyset, \emptyset\}$\\
        $\hat{\boldsymbol\xi}_2=\{\emptyset, \emptyset, \emptyset, \emptyset, \{1,2\}, \emptyset, \emptyset, \emptyset\}$\\
        $\hat{\boldsymbol\xi}_3=\{\emptyset, \emptyset, \emptyset, \emptyset, \{1,2\}, \emptyset, \{1\}, \emptyset\}$\\
        $\hat{\boldsymbol\xi}_4=\{\emptyset, \emptyset, \emptyset, \emptyset, \{1,2\}, \emptyset, \{1,\}, \{1,2\}\}$\\
        $\hat{\boldsymbol\xi}_5=\{\emptyset, \emptyset, \emptyset, \{1\}, \{1\}, \emptyset, \{1\}, \{1\}\}$\\
        \hline
        \end{tabular}
    \end{minipage}
\caption{
Regions of maximal ancestral state splits on the InvFels tree $\tau_1$.
}
\label{fig:max-anc-state}
\end{figure}

@@

The regions in Fig.~\ref{fig:max-anc-state} are analytically-derived regions of inconsistency in terms of probabilities of a character change along a branch for ``perfect'' data generated on the InvFels topology (Fig.~\ref{fig:farris-fels-top}) with $p_{w^*} = p_{y^*}$ (in terms of fidelities, $w^*=y^*$).
As the region of degeneracy in Fig.~\ref{fig:max-anc-state} gives the values of $x^*$ and $y^*$ where $\hat{w}$ is guaranteed to be one, we converge on a multifurcating topology in these cases.
%EM "letting $\emptyset$ be the maximal ancestral state split" sounds a little funny-- either it is or it isn't maximal.
It is easy to see that when $\emptyset$ is the maximal ancestral state split, we have the same log-likelihood for $\tau_1$ and $\tau_2$.
Moreover, if $w=1$, the internal branch becomes zero-length and the two topologies are indistinguishable.
Let $\mathcal{T}_0$ be such that, for $t^*=\{x^*,y^*,x^*,y^*,y^*\}$, $t^*\in\mathcal{T}_0$ corresponds to $x^*$ and $y^*$ falling in the region in Fig.~\ref{fig:max-anc-state} where $\hat{\boldsymbol\xi}=\hat{\boldsymbol\xi}_1$.
We can see this results in the likelihood of both topologies being equal, i.e.,
\begin{align*}
\max_{t : t^*\in\mathcal{T}_0} & \ \ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi)\\
&= \max_{t: \boldsymbol\xi=\hat{\boldsymbol\xi}_1, w=1, \tau=\tau_1} \ \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, \{x_1,y_1,x_2,y_2,w\})\\
&= \max_{t: \boldsymbol\xi=\hat{\boldsymbol\xi}_1, w=1, \tau=\tau_2} \ \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, \{x_1,y_1,x_2,y_2,w\}).
\end{align*}
